---
title: "Machine Learning Course Project"
author: "BrennenP"
date: "May 29, 2017"
output: html_document
---
##Project

###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

###Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

##Getting and Loading the Data
```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)

set.seed(39564)

trainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

##Reading the Data

After the data is loaded, the csv files are read into data frames.
```{r}
trainRaw <- read.csv(url(trainUrl))
testRaw <- read.csv(url(testUrl))

dim(trainRaw)
dim(testRaw)
```
The training set contains 19622 observations and 160 variables. The test set contains 20 observations and the same 160 variables. The "classe" variable is the one of interest here.

##Cleaning the Data

Now, The data will be cleaned of all observations with missing and meaningless values.
```{r}
sum(complete.cases(trainRaw))

trainRaw <- trainRaw[, colSums(is.na(trainRaw)) == 0] 
testRaw <- testRaw[, colSums(is.na(testRaw)) == 0] 

classe <- trainRaw$classe
trainRemove <- grepl("^X|timestamp|window", names(trainRaw))
trainRaw <- trainRaw[, !trainRemove]
trainCleaned <- trainRaw[, sapply(trainRaw, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testRaw))
testRaw <- testRaw[, !testRemove]
testCleaned <- testRaw[, sapply(testRaw, is.numeric)]

dim(trainCleaned)
dim(testCleaned)
```
The cleaned training set contains 19622 observations and 53 variables, and the cleaned test set contains 20 observations and the same 53 variables.

##Splitting the Data

The cleaned training set can be split into apure training set and a validation set. The validation set will be used in future steps.
```{r}
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
testData <- trainCleaned[-inTrain, ]
```

##Data Modeling

The data will be fit to a predictive model for activity recognition using Random Forest algorithm because it automatically selects important variables and is robust to correlated covariates & outliers in general. We will use 5-fold cross validation when applying the algorithm.
```{r}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf
```
Then the validation set is used to test the performance of the model.
```{r}
predictRf <- predict(modelRf, testData)
confusionMatrix(testData$classe, predictRf)
accuracy <- postResample(predictRf, testData$classe)
accuracy
oose <- 1 - as.numeric(confusionMatrix(testData$classe, predictRf)$overall[1])
oose
```
The estimated accuracy of the model is 99.47% and the out-of-sample error is 0.53%

##Predicting for Test Data

Finally, the model is applied to the testing data.
```{r}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```

##Appendix of Figures

#Correlation Matrix
```{r}
corrPlot <- cor(trainData[, -length(names(trainData))])
corrplot(corrPlot, method="color")
```

#Decision Tree
```{r}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel)
```